---
title: "Getting Started with FluxSeparator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with FluxSeparator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**FluxSeparator** separates diffusive and ebullitive (bubble) methane fluxes
from continuous concentration measurements. It uses a running variance approach
to identify sudden concentration increases caused by ebullition, then calculates
diffusive flux from the remaining data using linear regression.

For methodological details, see
[Sø et al. (2024)]<link href="\doi{10.1029/2024JG008035}" rel = "canonical">.

## Installation

Install the development version from GitHub:

```r
install.packages("FluxSeparator")
```

## Loading example data

The package includes a small dataset from Lake Lyng with two pump cycles: one
with only diffusive flux and one with an ebullitive event.

```{r load-data}
library(FluxSeparator)

data(DIY_sensor_data)
str(DIY_sensor_data)
```

## Calculating ebullitive flux

The `ebullitive_flux()` function identifies bubble events using a running
variance threshold. When the running variance exceeds `runvar_cutoff`, the
data is flagged as an ebullitive event.

```{r ebullitive}
ebul_flux <- ebullitive_flux(DIY_sensor_data, show_plots = FALSE)
ebul_flux
```

### Key parameters

- **`runvar_cutoff`** (default: 0.5): Lower values detect more bubbles. Start
  high and decrease until you capture the events you see in the data.
- **`IndexSpan`** (default: 30): How many observations to include before/after
  each detected event.
- **`top_selection`**: Use `"last"` (default) or `"max"` to determine the peak
  concentration of each bubble.
- **`show_plots`**: Set to `TRUE` for interactive visual inspection.

## Calculating diffusive flux

The `diffusive_flux()` function first removes ebullitive events (unless
`look_for_bubbles = FALSE`), then fits a linear model to the remaining
concentration data to estimate the diffusive flux.

```{r diffusive}
diff_flux <- diffusive_flux(DIY_sensor_data,
                            cutoff_start_value = 5,
                            show_plots = FALSE)
diff_flux
```

### Key parameters

- **`cutoff_start_value`**: Maximum starting concentration for a cycle to be
  included. Use `Inf` (default) to include all cycles.
- **`remove_observations_prior`** (default: 200): Skip the first N observations
  to allow the chamber to equilibrate.
- **`number_of_observations_used`** (default: 400): Number of observations for
  the linear regression.
- **`look_for_bubbles`** (default: `TRUE`): Set to `FALSE` for gases where
  ebullition is unlikely (e.g., CO2).

## Unit conversion

Convert from ppm per hour to µmol m⁻² h⁻¹ using the ideal gas law:

```{r conversion}
ppm_to_umol(
  pressure = 101325,     # Pa (standard atmospheric)
  concentration = 10,    # ppm/hr
  volume = 0.01,         # m³
  temperature_C = 20,    # °C
  area = 0.05            # m²
)
```

## Choosing the right `runvar_cutoff`

The most important parameter is `runvar_cutoff`. To find the best value:

1. Run `ebullitive_flux()` with `show_plots = TRUE`
2. Inspect the running variance plots
3. Adjust `runvar_cutoff` until bubble events are correctly identified
4. Use the **same** `runvar_cutoff` for both `ebullitive_flux()` and
   `diffusive_flux()`

## Working with plots

Both functions can return plot objects as an attribute when `show_plots = TRUE`:

```{r plots, eval = FALSE}
result <- ebullitive_flux(DIY_sensor_data, show_plots = TRUE)

# Access individual plots
plots <- attr(result, "plots")
```

## Citing FluxSeparator

```{r citation}
citation("FluxSeparator")
```
